---
title: "Practical Machine Learning - Prediction Assignment Writeup"
author: "Liew Chee Hau"
date: "1/17/2021"
output: html_document
---

## Overview
The goal of your project is to predict the manner in which they did the exercise. The machine learning algorithm, which uses the classe variable in the training set, is applied to the 20 test cases available in the test data.

## Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

## Data loading and Cleaning
Load the required R packages
```{r, echo=TRUE}
library(knitr)
library(caret)
library(rpart)
library(rpart.plot)
library(rattle)
library(randomForest)
```

Read the datasets
```{r, echo=TRUE}
urltrain <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
urltest  <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training <- read.csv(url(urltrain))
testing  <- read.csv(url(urltest))
```

Create data partition with the training dataset
```{r, echo=TRUE}
inTrain  <- createDataPartition(training$classe, p=0.7, list=FALSE)
TrainSet <- training[inTrain, ]
TestSet  <- training[-inTrain, ]
dim(TrainSet)
dim(TestSet)
```

There are total 160 variables for both datasets. There are plenty of variables with NA. Near Zero Variance (NZV) removed these variables and the ID variables as well.

```{r, echo=TRUE}
NZV <- nearZeroVar(TrainSet)
TrainSet <- TrainSet[, -NZV]
TestSet  <- TestSet[, -NZV]
dim(TrainSet)
dim(TestSet)
```

Remove variables that are mostly NA
```{r, echo=TRUE}
AllNA <- sapply(TrainSet, function(x) mean(is.na(x))) > 0.95
TrainSet <- TrainSet[, AllNA==FALSE]
TestSet  <- TestSet[, AllNA==FALSE]
dim(TrainSet)
dim(TestSet)
```

Remove identification only variables (columns 1 to 5)
```{r, echo=TRUE}
TrainSet <- TrainSet[, -(1:5)]
TestSet  <- TestSet[, -(1:5)]
dim(TrainSet)
dim(TestSet)
```

After the data cleaning process, the total number of variables used for analysis is 54 only.

## Prediction Model Building
Three methods will be applied to the model the regressions of the Train dataset and the method with the highest accuracy will be applied to the Test dataset for the quiz predictions. The methods are: Random Forests, Decision Tree and Generalized Boosted Model.

A Confusion Matrix is plotted at the end of each analysis to better visualize the accuracy of the models.

### 1) Random Forest
Model fit
```{r, echo=TRUE}
set.seed(1234)
controlRF <- trainControl(method="cv", number=3, verboseIter=FALSE)
modFitRandForest <- train(classe ~ ., data=TrainSet, method="rf", trControl=controlRF)
modFitRandForest$finalModel
```

Prediction on Test dataset
```{r, echo=TRUE}
predictRandForest <- predict(modFitRandForest, newdata=TestSet)
confMatRandForest<-confusionMatrix(predictRandForest,as.factor(TestSet$classe))
confMatRandForest
```

Plot results
```{r, echo=TRUE}
plot(confMatRandForest$table, col = confMatRandForest$byClass, 
     main = paste("Random Forest - Accuracy =",  round(confMatRandForest$overall['Accuracy'], 4)))
```

### 2) Decision Trees
Model Fit
```{r, echo=TRUE}
set.seed(1234)
modFitDecTree <- rpart(classe ~ ., data=TrainSet, method="class")
```

Prediction on Test dataset
```{r, echo=TRUE}
predictDecTree <- predict(modFitDecTree, newdata=TestSet, type="class")
confMatDecTree <- confusionMatrix(predictDecTree, as.factor(TestSet$classe))
confMatDecTree
```

Plot results
```{r, echo=TRUE}
plot(confMatDecTree$table, col = confMatDecTree$byClass, 
     main = paste("Decision Tree - Accuracy =",
                  round(confMatDecTree$overall['Accuracy'], 4)))
```


### 3) Generalized Boosted Model
Model Fit
```{r, echo=TRUE}
set.seed(1234)
controlGBM <- trainControl(method = "repeatedcv", number = 5, repeats = 1)
modFitGBM  <- train(classe ~ ., data=TrainSet, method = "gbm", trControl = controlGBM, verbose = FALSE)
modFitGBM$finalModel
```

Prediction on Test dataset
```{r, echo=TRUE}
predictGBM <- predict(modFitGBM, newdata=TestSet)
confMatGBM <- confusionMatrix(predictGBM, as.factor(TestSet$classe))
confMatGBM
```

Plot results
```{r, echo=TRUE}
plot(confMatGBM$table, col = confMatGBM$byClass, 
     main = paste("GBM - Accuracy =", round(confMatGBM$overall['Accuracy'], 4)))
```

## Conclusion
The accuracy of the 3 regression modeling methods above are:
1.Random Forest:0.9976,
2.Decision Tree:0.7366,
3.GBM : 0.9871

From the above results, the Random Forest model has the best accuracy and hence will be used to predict the 20 quiz results.

```{r, echo=TRUE}
predictTEST <- predict(modFitRandForest, newdata=testing)
predictTEST
```

